Komols Complexity, Coresets, and Sketches in Machine Learning

### 2019-01-03 komlos.tex
fixed streaming

The bottom layers will be those of height $h=0$ up to $h=H'$ where $H'$ is set adaptively in a way that $H'$ is the maximum integer such that for all $h \leq H'$, 
\begin{equation} \label{eq:Htag_nh}
 \forall h \leq H' , \ \ n_h \geq c_1 k \log^2(n) \log(1/(c_2 \delta \log(n)))
\end{equation}
for some sufficiently large constant $c_1$ and small constant $c_2$ to be determined later.

According to Lemma~\ref{lem:compactor} we have for $ h\leq H'$ that for some appropriate constant $c_1$ in  \eqref{eq:Htag_nh},
$$ \Pr[n_{h+1} \geq (1+1/\log(n))n_{h}/2] \leq \exp\left( -\frac{n_{h}}{k_h \log^2(n)}  \right) \leq c_2\delta / \log(n) $$
We can now union bound over all 
$$h \leq \log_2\left( n / \left( c_1 k \log^2(n) \log(1/(c_2 \delta\log(n))) \right)\right) + 2 \leq \log_2(n)$$
and conclude that w.p.\ at least $1-\delta/2$
\begin{equation} \label{eq:Htag}
H' \leq \log_2\left( n / \left( c_1 k \log^2(n) \log(1/(c_2 \delta \log(n))) \right)\right) + 2 \leq \log_2(n)
\end{equation}
and
\begin{equation} \label{eq:nh small}
\forall h \leq H', \ \ n_{h} \leq 3 \cdot n/2^{h}
\end{equation}

Let's proceed to bound the error of layer $h$. Since this error is a sum of i.i.d.\ bounded variables of mean zero we can use Chernoff- Hoeffding's inequality and obtain that
$$ \Pr\left[   \left| 2^h \sum_{j=1}^{\floor{n_h/k_h} } K_h Y_{ij} \right| \geq \eps n \right] \leq \exp \left( O\left( -\frac{ \eps^2 n^2}{ 2^{2h} \sum_{j=1}^{\floor{n_h/k_h} } K_h^2}   \right)\right) =$$
$$\exp \left( -O\left(\frac{\eps^2 n^2}{n_h 2^{2h} K_h^2/k_h  } \right) \right) 
$$
We now make use of the the bound on $n_h$ in Equation~\eqref{eq:nh small}. Notice that the random events determining the length $n_h$ do not affect the realization of the random variables used in layer $h$, so we can indeed use the bound on $n_h$ for controlling the error at level $h$. We get that since $n_h \leq 3n/2^h$
$$ \Pr\left[   \left| \text{Err}_h(q) \right| \geq \eps n \right] \leq 
\exp \left( -O\left(\frac{\eps^2 n}{ 2^h K_h^2/k_h  } \right) \right) 
$$
In particular, for $\eps_h = \sqrt{\frac{c 2^h K_h^2 \log((H'-h+1)^2/\delta)}{k_h n}}$ for some sufficiently large constant $c$ it holds that
$$ \Pr\left[   \left| \text{Err}_h(q) \right| \geq \eps n \right] \leq \delta/4(H'-h+1)^2$$
and a union bound ensures that with probability at least $1-\delta$, in addition to Equations~\eqref{eq:Htag} and~\eqref{eq:nh small} we have for all $h \leq H'$ that
$$\left| \text{Err}_h(q) \right| \leq \eps_h n$$
and 
$$\left| \sum_{h=0}^{H'} \text{Err}_h(q) \right| \leq \left(\sum_{h=0}^{H'} \eps_h\right)n $$

Let's proceed to bound the error on the top layers $H'+1$ to $H$. In these top layers we use a deterministic algorithm guaranteeing the stream is cut at least times 2 from layer to layer. We assign all top layers a budget of $k_h=k$ for the compactors. Since $2^{H+1} < n/k$ this gives a bound of 
$$
H-H' \leq \log_2(n/k2^{H'}) = O\left( \log\left( c_1 \log^2(n) \log(1/(c_2 \delta \log(n))) \right)\right) = 
$$
$$
O\left( \log \log(n/\delta) \right)
$$
leading to a bound of 
$$ \left| \sum_{h=H'+1}^H \text{Err}_h(q) \right| \leq O\left( \log \log(n/\delta) (K_k/k) n \right) $$
and an overall bound of
\begin{equation}
|\text{Err}(q)| \leq n \cdot O\left( \log \log(n/\delta)(K_k/k) + \sum_{h=0}^{H'} \eps_h \right)
\end{equation}

@@

We set $k_h = \max\{2, \ceil{(2/3)^{H'-h}k}\}$. Since $k_{H'} = k$ by using the definition of $H'$ in Equation~\eqref{eq:Htag} we see that
\begin{align*}
\eps_h = & O\left( \frac{K_k}{k} \sqrt{\frac{2^h K_h^2 k^2 \log((H'-h+1)^2/\delta)}{k_h K_k^2 n}} \right) = & n=\Omega(2^{H'}k\log(H'/\delta)) \\
  	   &  O\left( \frac{K_k}{k} \sqrt{\frac{2^h K_h^2 k }{k_h K_k^2 2^{H'}}} \right)  \\
    	   &  O\left( \frac{K_k}{k} \sqrt{\frac{K_h^2 k }{k_h K_k^2 2^{H'-h}}} \right)
\end{align*}
We move to control the sum over $\eps_h$ by bounding the multiple of $K_k/k$. For $k_h \geq d$
$$
\frac{K_h^2 k}{2^{H'-h} K_k^2 k_{h}} \leq \frac{d^2 k}{2^{H'-h} d^2 (2/3)^{H'-h}k} =
(3/4)^{H'-h}
$$
For $k_h \leq d$ we must have $h \leq H''$ with $(2/3)^{H'-H''} \leq d/k$, meaning that $2^{H'-H''} \geq k/d$. For such $h$ we get
\begin{align*}
 \frac{K_h^2 k}{2^{H'-h} K_k^2 k_{h}} = &  O\left( \frac{ k_h k}{2^{H'-h} d^2 } \right) =  \\
    	   &  O\left( \frac{ k_h k}{2^{H''-h} kd } \right) = & k_h \leq d\\
	   &  O\left( (1/2)^{H''-h}   \right) 
\end{align*}

@@

We are now ready for the case where $K_k = d\sqrt{k}$. We assume that $\sqrt{k} > d$ as otherwise we can result to a random sample. Here we set $H'' = H' - \ceil{\log_2(k/d^2)}$. For $h \geq H''$ we set $k_h=k$ and for $k \leq H''$ we set $k_h=2$. For $k \leq H''$ we have
\begin{align*}
\eps_h = & O\left( \sqrt{\frac{2^h K_h^2 \log((H'-h+1)^2/\delta)}{k_h n}} \right) = & n=\Omega(2^{H'}k\log(H'/\delta)), \ k_h=k=2 \\
  	   &  O\left( \sqrt{2^{h-H'}} \right)  = & 2^{H'-H''} \geq k/d^2 \\
    	   &  O\left(\frac{\sqrt{k}}{d} \sqrt{2^{h-H''}} \right)  
\end{align*}
and 
$$ \sum_{h=0}^{H''} \eps_h= O(\sqrt{k}/d) $$
\end{proof}
For $H'' \leq h \leq H'$ we use the deterministic algorithm and its error bound of
$$\sum_{h=H''}^{H'} |\text{Err}_h(q)| \leq O(\log(k/d^2) d/\sqrt{k})$$
Summing the errors of all layers we obtain a bound of
$$|\text{Err}(q)| \leq \frac{d}{\sqrt{k}} \cdot O(\log(k/d^2) + \log\log(n/\delta))$$
The memory requirement is $k \cdot O(\log(k/d^2) + \log\log(n/\delta))$ hence by setting 
$$k \approx \frac{d^2\left(\log^2(1/\eps)+ \log^2\log(n/\delta) \right)}{\eps^2}$$ 
we get a bound of $\eps n$ for the error and a memory usage of 
$$ O\left( \frac{d^2\left( \log^3(1/\eps) + \log^3\log(n/\delta) \right)}{\eps^2}\right) $$

>>>>

The bottom layers will be those of height $h=0$ up to $h=H'$ where $H'$ is set adaptively in a way that $H'$ is the maximum integer such that for all $h \leq H'$, 
\begin{equation} \label{eq:Htag_nh}
 \forall h \leq H' , \ \ n_h \geq c_1 k \log^2(n) \log(\log(n)/(c_2 \delta ))
\end{equation}
for some sufficiently large constant $c_1$ and small constant $c_2$ to be determined later.

According to Lemma~\ref{lem:compactor} we have for $ h\leq H'$ that for some appropriate constant $c_1$ in  \eqref{eq:Htag_nh},
$$ \Pr[n_{h+1} \geq (1+1/\log(n))n_{h}/2] \leq \exp\left( -\frac{n_{h}}{k_h \log^2(n)}  \right) \leq c_2\delta/  \log(n) $$
We can now union bound over all 
$$h \leq \log_2\left( n / \left( c_1 k \log^2(n) \log(1/(c_2 \delta\log(n))) \right)\right) + 2 \leq \log_2(n)$$
and conclude that w.p.\ at least $1-\delta/2$
\begin{equation} \label{eq:Htag}
H' \leq \log_2\left( n / \left( c_1 k \log^2(n) \log(1/(c_2 \delta \log(n))) \right)\right) + 2 \leq \log_2(n)
\end{equation}
and
\begin{equation} \label{eq:nh small}
\forall h \leq H', \ \ n_{h} \leq 3 \cdot n/2^{h}
\end{equation}

Let's proceed to bound the error of layer $h$. Since this error is a sum of i.i.d.\ bounded variables of mean zero we can use Chernoff- Hoeffding's inequality and obtain that
$$ \Pr\left[   \left| \text{Err}_h(q) \right| \geq \eps n \right] =$$
$$ \Pr\left[   \left| 2^h \sum_{j=1}^{\floor{n_h/k_h} } K_h Y_{ij} \right| \geq \eps n \right] \leq \exp \left( O\left( -\frac{ \eps^2 n^2}{ 2^{2h} \sum_{j=1}^{\floor{n_h/k_h} } K_h^2}   \right)\right) =$$
$$\exp \left( -O\left(\frac{\eps^2 n^2}{n_h 2^{2h} K_h^2/k_h  } \right) \right) 
$$
We now make use of the the bound on $n_h$ in Equation~\eqref{eq:nh small}. Notice that the random events determining the length $n_h$ do not affect the realization of the random variables used in layer $h$, so we can indeed use the bound on $n_h$ for controlling the error at level $h$. We get that since $n_h \leq 3n/2^h$
$$ \Pr\left[   \left| \text{Err}_h(q) \right| \geq \eps n \right] \leq 
\exp \left( -O\left(\frac{\eps^2 n}{ 2^h K_h^2/k_h  } \right) \right) 
$$
In particular, for $\eps_h = \sqrt{\frac{c 2^h K_h^2 \log((H'-h+1)^2/\delta)}{k_h n}}$ for some sufficiently large constant $c$ it holds that
$$ \Pr\left[   \left| \text{Err}_h(q) \right| \geq \eps n \right] \leq \delta/4(H'-h+1)^2$$
and a union bound ensures that with probability at least $1-\delta$, in addition to Equations~\eqref{eq:Htag} and~\eqref{eq:nh small} we have for all $h \leq H'$ that
$$\left| \text{Err}_h(q) \right| \leq \eps_h n$$
and 
$$\left| \sum_{h=0}^{H'} \text{Err}_h(q) \right| \leq \left(\sum_{h=0}^{H'} \eps_h\right)n $$

Let's proceed to bound the error on the top layers $H'+1$ to $H$. In these top layers we use a deterministic algorithm guaranteeing the stream is cut at least times 2 from layer to layer. We assign all top layers a budget of $k_h=k$ for the compactors. Since $2^{H+1} < n/k$ this gives a bound of 
$$
H-H' \leq \log_2(n/k2^{H'}) = O\left( \log\left( c_1 \log^2(n) \log(1/(c_2 \delta \log(n))) \right)\right) = 
$$
$$
O\left( \log \log(n/\delta) \right)
$$
leading to a bound of 
$$ \left| \sum_{h=H'+1}^H \text{Err}_h(q) \right| \leq O\left( \log \log(n/\delta) (K_k/k) n \right) $$
and an overall bound of
\begin{equation} \label{eq:err total}
|\text{Err}(q)| \leq n \cdot O\left( \log \log(n/\delta)(K_k/k) + \sum_{h=0}^{H'} \eps_h \right)
\end{equation}

@@

We set $k_h = \max\{2, \ceil{(2/3)^{H'-h}k}\}$. By using the definition of $H'$ in Equation~\eqref{eq:Htag} we see that
\begin{align*}
\eps_h = & O\left( \frac{K_k}{k} \sqrt{\frac{2^h K_h^2 k^2 \log((H'-h+1)^2/\delta)}{k_h K_k^2 n}} \right) = & n=\Omega(2^{H'}k\log(H'/\delta)) \\
  	   &  O\left( \frac{K_k}{k} \sqrt{\frac{2^h K_h^2 k }{k_h K_k^2 2^{H'}}} \right) = \\
    	   &  O\left( \frac{K_k}{k} \sqrt{\frac{K_h^2 k }{k_h K_k^2 2^{H'-h}}} \right)
\end{align*}
We move to control the sum over $\eps_h$ by bounding the multiple of $K_k/k$. For $k_h \geq d$
$$
\frac{K_h^2 k}{2^{H'-h} K_k^2 k_{h}} \leq \frac{d^2 k}{2^{H'-h} d^2 (2/3)^{H'-h}k} =
(3/4)^{H'-h}
$$
For $k_h \leq d$ we must have $h \leq H''$ with $(2/3)^{H'-H''} \leq d/k$, meaning that $2^{H'-H''} \geq k/d$. For such $h$ we get
\begin{align*}
 \frac{K_h^2 k}{2^{H'-h} K_k^2 k_{h}} = &  O\left( \frac{ k_h k}{2^{H'-h} d^2 } \right) =  \\
    	   &  O\left( \frac{ k_h k}{2^{H''-h} kd } \right) = & k_h \leq d\\
	   &  O\left( (1/2)^{H''-h}   \right) 
\end{align*}

@@

We are now ready for the case where $K_k = \min\{k, d\sqrt{k}\}$. We set $k_h=2$, resulting in $K_h=2$, for all $h \leq H'$ and obtain
\begin{align*}
\eps_h = & O\left( \sqrt{\frac{2^h K_h^2 \log((H'-h+1)^2/\delta)}{k_h n}} \right) = & n=\Omega(2^{H'}k\log(H'/\delta)), \ K_h^2/k_h=O(1) \\
  	   &  O\left( \sqrt{2^{h-H'}/k} \right) 
\end{align*}
and 
$$ \sum_h \eps_h = O\left(1/\sqrt{k}\right) $$
Plugging this to the overall error in Equation~\eqref{eq:err total} we get
$$ \text{Err}(q) = n \cdot O\left( \frac{\log \log(n/\delta)d + 1}{\sqrt{k}}  \right) =  n \cdot O\left( \frac{\log \log(n/\delta)d }{\sqrt{k}}  \right) $$

Since layers of $k_h=2$ can all be implemented in constant memory the memory requirement is dominated by the top layers and is in the order of $k \cdot \log\log(n/\delta)$, hence by setting 
$$k \approx \frac{d^2 \log^2\log(n/\delta)}{\eps^2}$$ 
we get a bound of $\eps n$ for the error and a memory usage of 
$$ O\left( \frac{d^2 \log^3\log(n/\delta)}{\eps^2}\right) $$
\end{proof}
